---
title: "Pirro Nord 13 - Reproducible Research"
subtitle: "Getting and cleaning data"
author: "Domenico Giusti"
output: html_document
  toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading data: RPostgreSQL connection and data.frame reading

```{r, echo=FALSE}
# Load required library
# https://code.google.com/p/rpostgresql/
library(RPostgreSQL)
```

Set first a connection with the PostgreSQL database

```{r}
# Establish connection to PoststgreSQL using RPostgreSQL
drv <- dbDriver("PostgreSQL")
con <- dbConnect(drv, dbname="pirronord",host="localhost",port=5432,user="dncgst",password="gnewpost")
# Check connection
dbListConnections(drv)
dbGetInfo(drv)
summary(con)
```

List tables of the database and the columns of a specific table

```{r}
# List tables
dbListTables(con)
# List column of a table
dbListFields(con,"record")
```

List views of the database

```{r}
# List views
# https://groups.google.com/d/topic/rpostgresql-dev/NgjFsIk8pCY/discussion
# dbSendQuery(con, statement, ...) submits one statement to the database
## Submits a statement
rs <- dbSendQuery(con, "SELECT viewname from pg_catalog.pg_views where viewowner = 'dncgst'")
## Fetch all elements from the result set
fetch(rs,n=-1)
## Or submit and execute the query directly
dbGetQuery(con, "SELECT viewname from pg_catalog.pg_views where viewowner = 'dncgst'")
```

Read a PostgreSQL table, or a subset of it, as a data.frame object (no spatial attribute). The geom is read as character!

```{r}
# Check if table exists. It works only with tables, no views!
dbExistsTable(con, "record")
# Read the complete db table in a data.frame
db_record <- dbReadTable(con,"record")
# Dataframe structure
class(db_record)
str(db_record)
```

Read in a data.frame from a query. It is a workaround to read a view

```{r}
# Read table lithic joining table record
db_lithic <- dbGetQuery(con, "SELECT DISTINCT ON (record.code) lithic.id, lithic.code, record.layer, record.us,	record.q,	record.n,	record.type, lithic.lithic_object, lithic.raw_material, lithic.raw_material_type, lithic.raw_material_code,	lithic.integrity,	lithic.length, lithic.width, lithic.thickness,	lithic.cortex, lithic.neocortex, lithic.patina, lithic.fracture, lithic.edge_ridge_rounding, record.date FROM public.lithic LEFT JOIN public.record ON lithic.code = record.code")
```

Close the database connection, once the required table are read

```{r}
# Close PostgreSQL connection 
dbDisconnect(con)
dbUnloadDriver(drv)
```

## Reading spatial data: readOGR a PostgreSQL/PostGIS relation in a Spatial*Dataframe

```{r, echo=FALSE}
# Load required library
# http://cran.r-project.org/web/packages/rgdal/
library(sp)
library(rgdal)
library(maptools)
```

Read a PostgreSQL/PostGIS view with _point_ geom using readOGR

```{r}
ogrDrivers()
# List the layers
ogrListLayers(dsn="PG:host=localhost port=5432 dbname=pirronord")
# Read the db view in a dataframe of class SpatialPointsDataframe
## view_sp_record: all coordinated records
sp_record <- readOGR(dsn="PG:host=localhost port=5432 dbname=pirronord", "view_sp_record")
class(sp_record)
bbox(sp_record)
proj4string(sp_record)
coordinates(sp_record)
names(sp_record)
str(sp_record)
## view_sp_lithic: all coordinated lithics with taphonomic attributes
sp_lithic <- readOGR(dsn="PG:host=localhost port=5432 dbname=pirronord", "view_sp_lithic")
spL <- sp_lithic[,c(1,3,17)]
## view_sp_fauna: all coordinated fossils with taphonomic attributes
sp_fauna <- readOGR(dsn="PG:host=localhost port=5432 dbname=pirronord", "view_sp_fauna")
spF <- sp_fauna[,c(1,3,16,17,18,20)]
## view_sp_micro: A.ruffoi
sp_micro <- readOGR(dsn="PG:host=localhost port=5432 dbname=pirronord", "view_micro")
```

Note: It will drop null geometries! Thus removing rows without xyx coordinates and rescaling the original bbox! Use dropNULLGeometries = FALSE will cause the matrix to be a simple data.frame without spatial attributes...

```{r}
# Combine the Spatial*Dataframe
## Bind A.ruffoi with the coordinated records
foo <- sp_record[,c(3,7)] #subset column: layer,type
bar <- sp_micro[,c(3,7)]
spX <- spRbind(foo, bar)
str(spX)
```

## Cleaning data

Check for duplicate

```{r}
# Check for duplication. Given a set of logical vectors, is at least one of the values true?
any(duplicated(sp_record$code))
# If TRUE, remove duplicated points
sp_record <- unique(sp_record)
```

Remove NA and NaN

```{r}

```

## Subsetting data

Subset records from SU A, B, C and D. Exclude SU E in order to work with a bounding box more close to the excavated area (the selected working window of the point processes)

```{r}
# Subset lithics from SU A,B,C,D
db_lithic <- subset(db_lithic, db_lithic$layer %in% c("A","B","C","D"))
##sp_lithic <- subset(sp_lithic, sp_lithic$layer %in% c("A","B","C","D"))
# Subset records from SU A,B,C,D
db_record <- subset(db_record, db_record$layer %in% c("A","B","C","D"))
spX <- subset(spX, spX$layer %in% c("A","B","C","D"))
```

Subset records by type: lithics, fossils, micro

```{r}
# Subset records not like DEl, Shell, Coprolite
db_record <- subset(db_record, !db_record$type %in% c("DEL","Coprolite","Shell"))
##sp_record <- subset(sp_record, !sp_record$type %in% c("DEL","Coprolite","Shell"))
```

```{r}
Xc <- spX[which(spX$layer == "C" & spX$type %in% c("Bone","Tooth","Lithic","Micro")),]
Xd <- spX[which(spX$layer == "D" & spX$type %in% c("Bone","Tooth","Lithic","Micro")),]
```

## Convert spatial data type: convert SpatialPointsDataframe to 'ppp' objects

```{r, echo=FALSE}
# Load libraries for point pattern analysis
# http://cran.r-project.org/web/views/Spatial.html
library(spatstat)
```

There are a number of packages for R which implement different functions for the analysis of spatial point patterns. These packages use different data structures to store the information of a point pattern. Package *maptools* offer some function to convert between _ppp_ objetcs ad _sp_ classes.

### Window

Set the bounding box

```{r}
# Bounding box of all records
bbox(sp_record)
```

If the natural window for the point pattern is not a rectangle, then you need to use a command like > ppp(x, y, window = W)

```{r}
# Define the window as excavated area
shp <- readShapePoly("/home/dncgst/project/pirronord_jas-reports/data/window/area_c.shp", proj4string=CRS(as.character(NA))) #import shp as SpatialPolygonsDataframe
shp <- as(shp, "SpatialPolygons") #change to SpatialPolygons
proj4string(shp) #check reference system = NA
bbox(shp) #check the extension of the bounding box
c <- as(shp, "owin") # convert to owin object
c
plot(c)

shp <- readShapePoly("/home/dncgst/project/pirronord_jas-reports/data/window/area_d.shp", proj4string=CRS(as.character(NA))) #import shp as SpatialPolygonsDataframe
shp <- as(shp, "SpatialPolygons") #change to SpatialPolygons
proj4string(shp) #check reference system = NA
bbox(shp) #check the extension of the bounding box
d <- as(shp, "owin") # convert to owin object
d
plot(d)
```

### Multitype point pattern

Extrude the z coordinate and use only xy coordinates for 2D analysis! Convert SpatialPointsDataframe in 'ppp' objects.

```{r}
# Convert the SpatialPointdataFrame to ppp objects
## Fossil and lithic record
X <- spCbind(Xd, Xd$coords.x3)
names(X)[3] <- "z"
record.d <- as(X[,c(2,3)], "ppp")
record.d$window <- d
unitname(record.d) <- "meter"
marks(record.d)
levels(record.d$marks$type)
record.d$marks$type <- droplevels(record.d$marks$type)
levels(record.d$marks$type) <- c("Fauna", "Lithic", "Fauna", "A.ruffoi")

###Independent random displacements to each point in a point pattern. Micromammals from the same sieve share the same point (center of 1sq.m unit). Scale of perturbation within 0.5m radius
record.d[which(record.d$marks$type == "A.ruffoi")] <- rjitter(record.d[which(record.d$marks$type == "A.ruffoi")], 0.5, retry=TRUE)
```
