---
title: "Pirro Nord 13 - Reproducible Research"
subtitle: "Point pattern analysis: post-depositional processes"
author: "Domenico Giusti"
output: html_document
  toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
# Load libraries for point pattern analysis
# http://cran.r-project.org/web/views/Spatial.html
library(spatstat)
```

Research questions:

> Strategy for analysis of multitype point pattern: conditional distribution P(M|X): regard the locations X as fixed or given, and study only the types M attached to these locations.

> In a conditional analysis [of a spatial case-control study] the [...] locations are treated as fixed covariates, and the desease status is treated as a random variable. [...] The simplest null model is that the desease status of each person is random, independently of other people, with a constant risk of desease (called 'random labelling')

## Distribution of patinae on the lithic assemblage

```{r}
# Getting and cleaning data
summary(Lc)

Xp.d <- as(Ld[,3], "ppp")
Xp.d$window <- d
unitname(Xp.d) <- "meter"
marks(Xp.d)
Xp.d <- subset(Xp.d, Xp.d$marks != "Brown")
Xp.d <- subset(Xp.d, Xp.d$marks != "White")
Xp.d$marks <- droplevels(Xp.d$marks)
```

```{r}
# Summary
plot(Xp.c)
summary(Xp.c)
```

```{r}
# Exploratory analysis of intensity
plot(density(split(Xp.c), sigma=bw.ppl(Xp.c), edge=TRUE, diggle=TRUE))
plot(density(split(Xp.c), sigma=bw.diggle(Xp.c), edge=TRUE, diggle=TRUE))
ProbU <- relrisk(Xp.c); plot(ProbU)
```

```{r}
# Correlation and spacing (Homogeneous process)
```

```{r}
# Test of randomness and independence
levels(Xp.c$marks) <- c("Absent","Fe-Mn","Fe-Mn")

E <- envelope(Xp.c, owin=c, fun=Kdif, i="Absent", nsim=19, simulate=expression(rlabel(Xp.c))); plot(E)
```

## Distribution of oxides on the lithic and the faunal assemblage

```{r}
summary(Fc)

Xo.c <- as(Fc[,c(3,4)], "ppp")
Xo.c$window <- c
unitname(Xo.c) <- "meter"
marks(Xo.c)
Xp.d <- subset(Xp.d, Xp.d$marks != "Brown")
Xp.d$marks <- droplevels(Xp.d$marks)
```

## Reference

Baddeley, A., Ruback, E., Turner, R., 2015. Spatial Point Patterns: Methodology and Applications with R. Chapman and Hall/CRC Press, London