---
title: "Pirro Nord 13 - Reproducible Research"
subtitle: "Exploratory Data Analysis: vertical distribution"
author: "Domenico Giusti"
output: html_document
  toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vertical distribution of finds

Asses for the normal distribution of the z coordinate before extruding it from subsequent point pattern analyses. We are interested in proceeding under the assumption that the data are 'nearly' normally distribuited on the third coordinate. Strongly (right) skewed distributions would occur in cases of non-uniform / non-simmetric / inhomogeneous vertical distribution of finds, thus requiring the analysis to account for a z covariate effect.

Many processes can be well approximated by the normal distribution. Testing the appropriateness of the normal assumption is a key step in many data analyses. 

```{r}
# Vertical distribution of 'sampled' finds in SU's C and D 

## Summary statistics
summary(Xc$coords.x3)

## Mean & Standard deviation
mean <- mean(Xc$coords.x3)
sd <- sd(Xc$coords.x3)

## Histogram
xlimC <- c(-3, 0)
xlimD <- c(-3.8, -.8)
hist(Xc$coords.x3, nclass=15, prob=TRUE, xlim=xlimC, xlab="z(meter)")
hist(Xd$coords.x3, nclass=15, prob=TRUE, xlim=xlimD, ylim=c(0,1.5), xlab="z(meter)")

## Fitting normal distribution with mean/sd of the data sample
curve(dnorm(x, mean(Xc$coords.x3), sd(Xc$coords.x3)), col="grey", lwd=2, add=TRUE)
curve(dnorm(x, mean=-1.535207114, sd=0.270189381), col="red", lwd=2, add=TRUE)
curve(dnorm(x, mean(Xd$coords.x3), sd(Xd$coords.x3)), col="grey", lwd=2, add=TRUE)
curve(dnorm(x, mean=-2.352142361, sd=0.334064143), col="red", lwd=2, add=TRUE)

### Maximum-likelihood fitting of univariate distributions
library(MASS)
fitdistr(Xc$coords.x3, "normal")
fitdistr(Xd$coords.x3, "normal")

#      mean         sd 
#-1.5352071  0.2701894
#fit <- rnorm(n=length(Xc$coords.x3), m=-1.5352071, sd=0.2701894)
#lines(density(fit))

shapiro.test(Xc$coords.x3)
shapiro.test(Xd$coords.x3)

ks.test(Xc$coords.x3, "pnorm")
ks.test(Xd$coords.x3, "pnorm")

## Histograms by type
hist(Xc$coords.x3[which(Xc$type %in% c("Bone","Tooth"))], nclass=15, prob=TRUE, xlim=xlimC, xlab="z(meter)")
hist(Xc$coords.x3[which(Xc$type == "Lithic")], nclass=15, prob=TRUE, xlim=xlimC, xlab="z(meter)")
hist(Xc$coords.x3[which(Xc$type == "Micro")], nclass=5, prob=TRUE, xlim=xlimC, xlab="z(meter)")

hist(Xd$coords.x3[which(Xd$type %in% c("Bone","Tooth"))], nclass=15, prob=TRUE, xlim=xlimD, xlab="z(meter)")
hist(Xd$coords.x3[which(Xd$type == "Lithic")], nclass=15, prob=TRUE, xlim=xlimD, xlab="z(meter)")
hist(Xd$coords.x3[which(Xd$type == "Micro")], nclass=10, prob=TRUE, xlim=xlimD, xlab="z(meter)")

## Density (kernel default:"gaussian"; bandwidth default:"nrd0")
### Bandwidth selectors for Gaussian kernels
bw.nrd0(Xc$coords.x3) # implements a rule-of-thumb for choosing the bandwidth of a Gaussian kernel density estimator. It defaults to 0.9 times the minimum of the standard deviation
bw.nrd(Xc$coords.x3) # is the more common variation given by Scott (1992), using factor 1.06
bw.bcv(Xc$coords.x3) #  implement biased cross-validation

lines(density(Xc$coords.x3, bw="nrd"), col="grey", lwd=1)
lines(density(Xc$coords.x3[which(Xc$type == "Bone")], bw="nrd"), col="#f8766d", lwd=1)
lines(density(Xc$coords.x3[which(Xc$type == "Tooth")], bw="nrd"), col="#f564e3", lwd=1)
lines(density(Xc$coords.x3[which(Xc$type == "Lithic")], bw="nrd"), col="#00ba38", lwd=1)
abline(v=mean(Xc$coords.x3[which(Xc$type == "Lithic")]), col="#00ba38", lwd=1, lty=2)
lines(density(Xc$coords.x3[which(Xc$type == "Micro")], bw="nrd0"), col="#00bfc4", lwd=1)
abline(v=mean(Xc$coords.x3[which(Xc$type == "Micro")]), col="#00bfc4", lwd=1, lty=2)

lines(density(Xd$coords.x3, bw="nrd"), col="grey", lwd=1)
lines(density(Xd$coords.x3[which(Xd$type == "Bone")], bw="nrd"), col="#f8766d", lwd=1)
lines(density(Xd$coords.x3[which(Xd$type == "Tooth")], bw="nrd"), col="#f564e3", lwd=1)
lines(density(Xd$coords.x3[which(Xd$type == "Lithic")], bw="nrd"), col="#00ba38", lwd=1)
abline(v=mean(Xd$coords.x3[which(Xd$type == "Lithic")]), col="#00ba38", lwd=1, lty=2)
lines(density(Xd$coords.x3[which(Xd$type == "Micro")], bw="nrd0"), col="#00bfc4", lwd=1)
abline(v=mean(Xd$coords.x3[which(Xd$type == "Micro")]), col="#00bfc4", lwd=1, lty=2)

## QQ plot
qqnorm(Xc$coords.x3); qqline(Xc$coords.x3, col=2)
qqnorm(Xc$coords.x3[which(Xc$type == "Lithic")]); qqline(Xc$coords.x3[which(Xc$type == "Lithic")], col=2)

qqnorm(Xd$coords.x3); qqline(Xd$coords.x3, col=2)
qqnorm(Xd$coords.x3[which(Xd$type == "Lithic")]); qqline(Xd$coords.x3[which(Xd$type == "Lithic")], col=2)

## Boxplot
boxplot(Xc$coords.x3)
boxplot(Xd$coords.x3)
```

### Vertical distribution of patinae on the lithic assemblage

```{r}
# Subset data (sample population of lithics)
spL <- spL[which(spL$layer %in% c("C", "D")),]
spL <- spL[complete.cases(spL$patina),]

summary(spL)

Lc <- spL[which(spL$layer == "C"),]
Ld <- spL[which(spL$layer == "D"),]

# Vertical distribution of taphonomic features (patinae) on the 'sampled' lithic assemblage
hist(spL$coords.x3, nclass=25, prob=TRUE, xlim=c(-3.5, 0), main="b", xlab="z(meter)")
lines(density(spL$coords.x3[which(spL$patina == "Absent")], bw="nrd"), col="#f8766d", lwd=1)
abline(v=mean(spL$coords.x3[which(spL$patina == "Absent")]), col="#f8766d", lwd=1, lty=2)
lines(density(spL$coords.x3[which(spL$patina == "Fe-Mn spotted")], bw="nrd"), col="#619cff", lwd=1)
abline(v=mean(spL$coords.x3[which(spL$patina == "Fe-Mn spotted")]), col="#619cff", lwd=1, lty=2)
lines(density(spL$coords.x3[which(spL$patina == "Fe-Mn covering")], bw="nrd"), col="#00ba38", lwd=1)
abline(v=mean(spL$coords.x3[which(spL$patina == "Fe-Mn covering")]), col="#00ba38", lwd=1, lty=2)
legend(x="topright", c("Absent","Fe-Mn spotted","Fe-Mn covering"), lty=1, lwd=1, col=c("#f8766d","#619cff","#00ba38"))

hist(Xc$coords.x3[which(Xc$type == "Lithic")], nclass=15, prob=TRUE, xlim=xlimC, xlab="z(meter)")
lines(density(Lc$coords.x3[which(Lc$patina == "Absent")], bw="nrd"), col="#f8766d", lwd=1)
abline(v=mean(Lc$coords.x3[which(Lc$patina == "Absent")]), col="#f8766d", lwd=1, lty=2)
lines(density(Lc$coords.x3[which(Lc$patina == "Fe-Mn covering")], bw="nrd"), col="#00ba38", lwd=1)
abline(v=mean(Lc$coords.x3[which(Lc$patina == "Fe-Mn covering")]), col="#00ba38", lwd=1, lty=2)
lines(density(Lc$coords.x3[which(Lc$patina == "Fe-Mn spotted")], bw="nrd"), col="#619cff", lwd=1)
abline(v=mean(Lc$coords.x3[which(Lc$patina == "Fe-Mn spotted")]), col="#619cff", lwd=1, lty=2)
legend(x="topright", c("Absent","Fe-Mn covering","Fe-Mn spotted"), lty=1, lwd=1, col=c("#f8766d","#00ba38","#619cff"))

hist(Xd$coords.x3[which(Xd$type == "Lithic")], nclass=15, prob=TRUE, xlim=xlimD, xlab="z(meter)")
lines(density(Ld$coords.x3[which(Ld$patina == "Absent")], bw="nrd"), col="#f8766d", lwd=1)
abline(v=mean(Ld$coords.x3[which(Ld$patina == "Absent")]), col="#f8766d", lwd=1, lty=2)
lines(density(Ld$coords.x3[which(Ld$patina == "Fe-Mn covering")], bw="nrd"), col="#00ba38", lwd=1)
abline(v=mean(Ld$coords.x3[which(Ld$patina == "Fe-Mn covering")]), col="#00ba38", lwd=1, lty=2)
lines(density(Ld$coords.x3[which(Ld$patina == "Fe-Mn spotted")], bw="nrd"), col="#619cff", lwd=1)
abline(v=mean(Ld$coords.x3[which(Ld$patina == "Fe-Mn spotted")]), col="#619cff", lwd=1, lty=2)
legend(x="topright", c("Absent","Fe-Mn covering","Fe-Mn spotted"), lty=1, lwd=1, col=c("#f8766d","#00ba38","#619cff"))
```

            patina  C
 Absent        :17  50%
 Brown         : 0  
 Fe-Mn covering: 4  
 Fe-Mn spotted :10  41%
 White         : 2  6%
 NA's          : 1  3%

            patina  D
 Absent        :40  48%
 Brown         : 1  1%
 Fe-Mn covering:14  
 Fe-Mn spotted :24  45%
 White         : 2  2%
 NA's          : 3  4%


### Vertical distribution of oxides on the faunal assemblage

```{r}
# Subset data (sample population of analized fossils)
Fc <- spF[which(spF$layer == "C"),]
Fc <- Fc[complete.cases(Fc$oxid),]

str(Fc)
summary(Fc)

levels(Fc$oxid)
levels(Fc$oxid) <- c("Low","Low","Low","Medium","Medium","Medium","High","High")

levels(Fc$abrasion)
levels(Fc$abrasion) <- c("Low","Low","Medium","High","High")

par(mfrow=c(1,3), cex=1)

# Vertical distribution of taphonomic features (oxides) on the 'sampled' faunal assemblage
hist(Fc$coords.x3, nclass=10, prob=TRUE, xlim=xlimC, ylim=c(0, 3), xlab="z(meter)", main="oxid")
lines(density(Fc$coords.x3[which(Fc$oxid == "Low")], bw="nrd"), col="#f8766d", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$oxid == "Low")]), col="#f8766d", lwd=1, lty=2)
lines(density(Fc$coords.x3[which(Fc$oxid == "Medium")], bw="nrd"), col="#619cff", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$oxid == "Medium")]), col="#619cff", lwd=1, lty=2)
lines(density(Fc$coords.x3[which(Fc$oxid == "High")], bw="nrd"), col="#00ba38", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$oxid == "High")]), col="#00ba38", lwd=1, lty=2)
#legend(x="topright", c("Low","Medium","High"), lty=1, lwd=1, col=c("#f8766d","#619cff","#00ba38"))

# Vertical distribution of taphonomic features (abrasion) on the 'sampled' faunal assemblage
hist(Fc$coords.x3, nclass=10, prob=TRUE, xlim=xlimC, ylim=c(0, 4), xlab="z(meter)", main="abrasion")
lines(density(Fc$coords.x3[which(Fc$abrasion == "Low")], bw="nrd"), col="#f8766d", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$abrasion == "Low")]), col="#f8766d", lwd=1, lty=2)
lines(density(Fc$coords.x3[which(Fc$abrasion == "Medium")], bw="nrd"), col="#619cff", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$abrasion == "Medium")]), col="#619cff", lwd=1, lty=2)
lines(density(Fc$coords.x3[which(Fc$abrasion == "High")], bw="nrd"), col="#00ba38", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$abrasion == "High")]), col="#00ba38", lwd=1, lty=2)
#legend(x="topright", c("Low","Medium","High"), lty=1, lwd=1, col=c("#f8766d","#619cff","#00ba38"))

# Vertical distribution of taphorecord on the 'sampled' faunal assemblage
hist(Fc$coords.x3, nclass=10, prob=TRUE, xlim=xlimC, ylim=c(0, 5), xlab="z(meter)", main="TR")
lines(density(Fc$coords.x3[which(Fc$taphorecord == "3")], bw="nrd"), col="#f8766d", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$taphorecord == "3")]), col="#f8766d", lwd=1, lty=2)
lines(density(Fc$coords.x3[which(Fc$taphorecord == "2")], bw="nrd"), col="#619cff", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$taphorecord == "2")]), col="#619cff", lwd=1, lty=2)
lines(density(Fc$coords.x3[which(Fc$taphorecord == "1")], bw="nrd"), col="#00ba38", lwd=1)
#abline(v=mean(Fc$coords.x3[which(Fc$taphorecord == "1")]), col="#00ba38", lwd=1, lty=2)
#legend(x="topright", c("3","2","1"), lty=1, lwd=1, col=c("#f8766d","#619cff","#00ba38"))
```

  oxid
   Low Medium   High
    19     43     73
